---
output: 
  html_document: 
    keep_md: yes
---
# Reproducible Research Course- Project 1 (activity.csv)
Author       : Ralston Fonseca 
Date         : 04 Sep 2018
Version      : 1.0
Description  : This is R markdown file for analysing "activity.csv" 


The data for this assignment was be downloaded from the course web site [Activity monitoring data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip) [52K]
Refer to README.md for more details.

## Loading and preprocessing the data

Following code was used to read the data from local working directory.
```{r}
activityDf <- read.csv("activity.csv", header = TRUE)
dim(activityDf)
```
It consists of *17568* observations and *3* variables.

There are roughly 13% values which are missing in *steps* variable

```{r}
mean(is.na(activityDf$steps))
```

## What is mean total number of steps taken per day?

The missing values are ignored for plotting the histogram and getting the mean and median

```{r}
activityDFOmittingNA <- subset(activityDf, complete.cases(activityDf)) 

#Get the mean
dailySteps <- tapply(activityDFOmittingNA$steps,activityDFOmittingNA$date,sum, na.rm = TRUE)

# Make a histogram of the total number of steps taken each day
hist(dailySteps, xlab = "Number of steps", ylab = "Number of days", 
     main = "Total number of steps taken each day", col = "aquamarine4")

# Mean
mean(dailySteps, na.rm = TRUE)

#Median
median(dailySteps, na.rm = TRUE)

```

## What is the average daily activity pattern?

The following steps takes into account the entire dataset.


1. The following graph depicts time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r}
# Full dataset used
avgStepsPerInterval <- tapply(activityDf$steps,activityDf$interval ,mean, na.rm = TRUE)

plot(names(avgStepsPerInterval), avgStepsPerInterval, type = "l", 
     xlab = "Interval (5 min identifier)", ylab = "Average number of steps", 
     main = "Average daily activity pattern", col = "aquamarine4")
```

2. As shown below the *835*, 5-minute interval, contains the maximum number of steps on average across all the days:

```{r}
avgStepsPerInterval[which.max(avgStepsPerInterval)]
```

## Imputing missing values

```{r}
# All dates have 288 entries
table(activityDf$date) 

activityDFMissingSteps <- subset(activityDf, is.na(activityDf$steps)) # dim 2304 3

# All 288 values of steps are missing for these dates
table(activityDFMissingSteps$date) 

# 8 values missing per interval
table(activityDFMissingSteps$interval) 

nmissing <- nrow(activityDFMissingSteps)

```
1) The total rows with missing values are: `r nmissing`

2) As shown above the entire set for that *date* is missing so I have decided to fill missing values by *interval*. Also the using *median* will skew the data on the lower side as there considerable values with 0 steps for higher intervals. Hence I have decided to use *mean of steps for an interval* to fill the missing values.

3) Creating a new dataset to fill the missing values. Values are rounded to lower integer value.

```{r}
# Create new dataset
activityFilledNA <- activityDf

# fill the missing values with mean of steps for an interval
for (i in activityFilledNA$interval) {
    activityFilledNA[which( activityFilledNA$interval == i & 
                                is.na(activityFilledNA$steps)),"steps"] <- 
        round(avgStepsPerInterval[names(avgStepsPerInterval) == i])
}
```

4) Plotting histogram and calculating mean and median using filled up values for missing data.

```{r}

dailySteps2 <- tapply(activityFilledNA$steps,activityFilledNA$date,sum, na.rm = TRUE)


# Make a histogram of the total number of steps taken each day
hist(dailySteps2, xlab = "Number of steps", ylab = "Number of days", 
     main = "Total number of steps taken each day", col = "aquamarine4")

# Calculate mean
mean(dailySteps2, na.rm = TRUE)

# Calculate median
median(dailySteps2, na.rm = TRUE)
```

As we see the *mean* and *median* values have reduced slightly but are still close based on the above strategy.

## Are there differences in activity patterns between weekdays and weekends?

Create extra column *weekday* to categorise the row as *weekday* or *weekend*.
Using *lattice* package to create the plot.

```{r}
library(lattice)

activityFilledNA$weekday <- ifelse(weekdays(as.Date(activityFilledNA$date)) %in% c("Saturday", "Sunday"), "weekend", "weekday")

avgSPI <-aggregate(steps ~ weekday + interval,data = activityFilledNA, FUN = mean )

xyplot(steps ~ interval | weekday, data = avgSPI, type = "l", layout = c(1, 2),
       xlab = "Interval (5 min identifier)", ylab = "Number of steps",
       main = "Average number of steps")
```
